Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-03-29 deprecated.tex
updates
reworking some bounds/notation

>>>>

\begin{align*}
    0 &\ge h_{\tau_1,\{x,y,y\}}(\tau_1,\{y, y, y\}) + L^{p}_{\tau_1,\tau_1}(a_{xy},b_{xy}) \\
      &\qquad - h_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-d_{xy}-2e_{xy}, 0\}) - L^{p,1}_{\tau_1,\tau_2}(d_{xy},e_{xy}) \\
%
      &\qquad = 2(1+x^2-y^2-x^2y^2)\log(1+x^2-y^2-x^2y^2) \\
      &\qquad + 2(1-x^2+y^2-x^2y^2)\log(1-x^2+y^2-x^2y^2) \\
      &\qquad + 2(1-x^2-y^2+x^2y^2)\log(1-x^2-y^2+x^2y^2) \\
      &\qquad + (1+x^2+y^2+4xy^2+x^2y^2)\log(1+x^2+y^2+4xy^2+x^2y^2) \\
      &\qquad + (1+x^2+y^2-4xy^2+x^2y^2)\log(1+x^2+y^2-4xy^2+x^2y^2) \\
%
      &\qquad + (2-a_{xy})\log(2-a_{xy})+a_{xy}\log(a_{xy})+(2-b_{xy})\log(2-b_{xy})+b_{xy}\log(b_{xy})+\log(2) \\
%
      &\qquad - 2(1+x^2-y^2-x^2y^2)\log(1+x'^2w'-y'^2w'-x'^2y'^2) \\
      &\qquad - 2(1-x^2+y^2-x^2y^2)\log(1-x'^2w'+y'^2w'-x'^2y'^2) \\
      &\qquad - (1-x^2-y^2+x^2y^2)\log(1+2x'y'-2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2) \\
      &\qquad - (1-x^2-y^2+x^2y^2)\log(1-2x'y'+2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2) \\
      &\qquad - (1+x^2+y^2+4xy^2+x^2y^2)\log(1+2x'y'+2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2) \\
      &\qquad - (1+x^2+y^2-4xy^2+x^2y^2)\log(1-2x'y'-2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2) \\
%
      &\qquad - 2\log(2)-(2-d_{xy}-2e_{xy})\log(2-d_{xy}-2e_{xy})-(d_{xy}+2e_{xy})\log(d_{xy}+2e_{xy})
\end{align*}
where
$$
a_{xy} = b_{xy} = \frac{4-2x^2-2y^2}{8},
$$
$$
d_{xy} = \frac{4-4x^2y^2}{8},
$$
$$
e_{xy} = \frac{2-4xy^2+2x^2y^2}{8}
$$
and
$$
x' = 1, \ y' = 1-d_{xy}-2e_{xy}, \ w' = 0.
$$

The \texttt{sage} function in listing~\ref{list:sage} will generate Fig.~\ref{fig:inconsistency-farris}.

\begin{table}
    \begin{center}
\begin{verbatim}
var('x,y')
a = (4-2*x^2-2*y^2)/8
b = (4-2*x^2-2*y^2)/8
c = (4-4*x^2*y^2)/8
d = (2-4*x*y^2+2*x^2*y^2)/8
xp = 1.
yp = 1-c-2*d
wp = 0.

inconsistency = 2*(1+x^2-y^2-x^2*y^2)*log(1+x^2-y^2-x^2*y^2) \
    + 2*(1-x^2+y^2-x^2*y^2)*log(1-x^2+y^2-x^2*y^2) \
    + 2*(1-x^2-y^2+x^2*y^2)*log(1-x^2-y^2+x^2*y^2) \
    + (1+x^2+y^2+4*x*y^2+x^2*y^2)*log(1+x^2+y^2+4*x*y^2+x^2*y^2) \
    + (1+x^2+y^2-4*x*y^2+x^2*y^2)*log(1+x^2+y^2-4*x*y^2+x^2*y^2) \
    + (2-a)*log(2-a)+a*log(a)+(2-b)*log(2-b)+b*log(b)+log(2.) \
    - 2*(1+x^2-y^2-x^2*y^2)*log(1+xp^2*wp-yp^2*wp-xp^2*yp^2) \
    - 2*(1-x^2+y^2-x^2*y^2)*log(1-xp^2*wp+yp^2*wp-xp^2*yp^2) \
    - (1-x^2-y^2+x^2*y^2)*log(1+2*xp*yp-2*xp*yp*wp-xp^2*wp-yp^2*wp+xp^2*yp^2) \
    - (1-x^2-y^2+x^2*y^2)*log(1-2*xp*yp+2*xp*yp*wp-xp^2*wp-yp^2*wp+xp^2*yp^2) \
    - (1+x^2+y^2+4*x*y^2+x^2*y^2)*log(1+2*xp*yp+2*xp*yp*wp+xp^2*wp+yp^2*wp+xp^2*yp^2) \
    - (1+x^2+y^2-4*x*y^2+x^2*y^2)*log(1-2*xp*yp-2*xp*yp*wp+xp^2*wp+yp^2*wp+xp^2*yp^2) \
    - 2*log(2.)-(2-c-2*d)*log(2-c-2*d)-(c+2*d)*log(c+2*d)

incons_plot = implicit_plot(
    inconsistency(x,y)==0,
    (x,0,1),
    (y,0,1),
    axes_labels=['$x$', '$y$'],
    title='Region of inconsistency for Farris-generating topology',
    color='black'
)
\end{verbatim}
    \end{center}
\caption{Sage code to generate Fig.~\ref{fig:inconsistency-farris}}
\label{list:sage}
\end{table}

To obtain an upper bound for the Farris partial likelihood---the terms in the top half of Table~\ref{tab:likelihoods}---we use the term $(1+x)^2(1+w)(1+y)^2$ for the $\{1,3\}$ split.
Since all cases have a $(1+x)^2$ term except $\{\{1\}, \{2\}, \{1,2\}, \{2,3\}\}$, and all cases except $\{\{3\}, \{1,2\}, \{2,3\}, \{1,2,3\}\}$ have a $(1+y)^2$, we can add and subtract the $\log$ of these terms weighted by each of their generating probabilities.
In other words, define
\begin{align*}
    a_{xy} &= P(\alignmentColumn=\{1\}|\tau=\tau_1,t=\{x,y,y\})
    +        P(\alignmentColumn=\{2\}|\tau=\tau_1,t=\{x,y,y\}) \\
    &\qquad + P(\alignmentColumn=\{1,2\}|\tau=\tau_1,t=\{x,y,y\})
    + P(\alignmentColumn=\{2,3\}|\tau=\tau_1,t=\{x,y,y\}),
\end{align*}
\begin{align*}
    b_{xy} &= P(\alignmentColumn=\{3\}|\tau=\tau_1,t=\{x,y,y\}) + P(\alignmentColumn=\{1,2\}|\tau=\tau_1,t=\{x,y,y\}) \\
           &\qquad + P(\alignmentColumn=\{2,3\}|\tau=\tau_1,t=\{x,y,y\}) + P(\alignmentColumn=\{1,2,3\}|\tau=\tau_1,t=\{x,y,y\})
\end{align*}
and
\begin{equation*}
    c_{xy} = P(\alignmentColumn=\{1,3\}|\tau=\tau_1,t=\{x,y,y\}).
\end{equation*}
Then the partial likelihood is bounded above by
\begin{align*}
    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_1,\{x',y',w'\}) \\
    &\qquad \le (2-a_{xy})\log(1+x')+a_{xy}\log(1-x')\\
    &\qquad \qquad +(2-b_{xy})\log(1+y')+b_{xy}\log(1-y')\\
    &\qquad \qquad +(1-c_{xy})\log(1+w')+c_{xy}\log(1-w') \\
    &\qquad \le (2-a_{xy})\log(2-a_{xy})+a_{xy}\log(a_{xy})\\
    &\qquad \qquad +(2-b_{xy})\log(2-b_{xy})+b_{xy}\log(b_{xy})\\
    &\qquad \qquad +(1-c_{xy})\log(1-c_{xy})+c_{xy}\log(c_{xy})+\log(2) \\
    &\qquad \triangleq L^{p,1}_{\tau_1,\tau_1}(a_{xy}, b_{xy})
\end{align*}
or
\begin{align*}
    & \ell^p_{\tau_1,\{x,y,y\}}(\fullAncestralStateCategories,\tau_1,\{x',y',w'\}) \\
    &\qquad \le (2-a_{xy}-2c_{xy})\log(1+x')+(a_{xy}+2c_{xy})\log(1-x')\\
    &\qquad \qquad +(2-b_{xy})\log(1+y')+b_{xy}\log(1-y')+\log(1+w')\\
    &\qquad \le (2-a_{xy}-2c_{xy})\log(2-a_{xy}-2c_{xy})+(a_{xy}+2c_{xy})\log(a_{xy}+2c_{xy})\\
    &\qquad \qquad +(2-b_{xy})\log(2-b_{xy})+b_{xy}\log(b_{xy})+\log(2) \\
    &\qquad \triangleq L^{p,2}_{\tau_1,\tau_1}(a_{xy}, b_{xy})
\end{align*}
where $a_{xy},b_{xy}$ and $c_{xy}$ are functions of the true generating branch parameters $\{x,y\}$.

\begin{table}
\centering
\begin{tabular}{|l|ll|}
\multicolumn{3}{c}{$P(\ancestralStateColumn=\ancestralSplit_k|\alignmentColumn=\siteSplit_j,\tau_1,t)$}\\
\hline
& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
    \hline
    $\siteSplit_j$    &$\emptyset$  00                              &$\{2\}$  01\\
    \hline
     $\emptyset$   0000&$(1+x)^2   (1+w)(1+y)^{2*}$          &$(1+x)^2   (1-w)(1-y)^2$\\
     $\{1\}$       1000&$(1+x)(1-x)(1+w)(1+y)^{2*}$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{2\}$       0100&$(1+x)(1-x)(1+w)(1+y)^{2*}$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{3\}$       0010&$(1+x)^2   (1+w)(1+y)(1-y)^*$        &$(1+x)^2   (1-w)(1+y)(1-y)$\\
     $\{1,2\}$     0011&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$&$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1,3\}$     0101&$(1+x)^2   (1+w)(1-y)^{2\ddagger}$   &$(1+x)^2   (1-w)(1+y)^{2\ddagger}$\\
     $\{2,3\}$     1001&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$&$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1,2,3\}$   0001&$(1+x)^2   (1+w)(1+y)(1-y)^*$        &$(1+x)^2   (1-w)(1+y)(1-y)$\\
    \hline
    \hline
    &$\{1\}$  10                           &$\{1,2\}$  11\\
    \hline
     $\emptyset$   0000&$(1-x)^2   (1-w)(1+y)^2$     &$(1-x)^2   (1+w)(1-y)^2$\\
     $\{1\}$       1000&$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{2\}$       0100&$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{3\}$       0010&$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
     $\{1,2\}$     0011&$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$\\
     $\{1,3\}$     0101&$(1-x)^2   (1-w)(1-y)^2$     &$(1-x)^2   (1+w)(1+y)^{2\ddagger}$\\
     $\{2,3\}$     1001&$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$\\
     $\{1,2,3\}$   0001&$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
    \hline
\multicolumn{3}{c}{$P(\ancestralStateColumn=\ancestralSplit_k|\alignmentColumn=\siteSplit_j,\tau_2,t)$}\\
\hline
& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
    \hline
    $\siteSplit_j$    &$\emptyset$  00                              &$\{2\}$  01\\
    \hline
     $\emptyset$   0000&$(1+x)^2   (1+w)(1+y)^{2*}$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1\}$       1000&$(1+x)(1-x)(1+w)(1+y)^{2\ddagger}$    &$(1-x)^2   (1-w)(1+y)(1-y)$\\
     $\{2\}$       0100&$(1+x)^2   (1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{3\}$       0010&$(1+x)(1-x)(1+w)(1+y)^{2\ddagger}$    &$(1+x)^2   (1-w)(1+y)(1-y)^{\ddagger}$\\
     $\{1,2\}$     0011&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)^2   (1-w)(1+y)^{2\ddagger}$\\
     $\{1,3\}$     0101&$(1+x)^2   (1+w)(1-y)^{2\ddagger}$    &$(1+x)(1-x)(1-w)(1+y)(1-y)^{\ddagger}$\\
     $\{2,3\}$     1001&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$&$(1-x)^2   (1-w)(1+y)^{2\ddagger}$\\
     $\{1,2,3\}$   0001&$(1+x)^2   (1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)(1-x)(1-w)(1+y)^{2\ddagger}$\\
    \hline
    \hline
    &$\{1\}$  10                           &$\{1,2\}$  11\\
    \hline
     $\emptyset$   0000&$(1+x)(1-x)(1-w)(1+y)(1-y)$             &$(1-x)^2   (1+w)(1-y)^2$\\
     $\{1\}$       1000&$(1+x)^2   (1-w)(1+y)(1-y)^{\ddagger}$  &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{2\}$       0100&$(1+x)(1-x)(1-w)(1+y)^{2\ddagger}$      &$(1-x)^2   (1+w)(1+y)(1-y)$\\
     $\{3\}$       0010&$(1-x)^2   (1-w)(1+y)(1-y)$             &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{1,2\}$     0011&$(1-x)^2   (1-w)(1-y)^2$                &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$\\
     $\{1,3\}$     0101&$(1+x)(1-x)(1-w)(1+y)(1-y)^{\ddagger}$  &$(1-x)^2   (1+w)(1+y)^{2\ddagger}$\\
     $\{2,3\}$     1001&$(1+x)^2   (1-w)(1-y)^{2\ddagger}$      &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$\\
     $\{1,2,3\}$   0001&$(1+x)(1-x)(1-w)(1-y)^2$                &$(1-x)^2   (1+w)(1+y)(1-y)$\\
\hline
\end{tabular}
\caption{Likelihood calculations for all site patterns and internal states of Farris topology.
Maxima determined row-wise (i.e., by site pattern).
All values multiplied by $1/32$.
Key: $^*$ unique maximum value corresponding to unique internal state; $^\dagger$ unique maximum value corresponding to multiple internal states; $^\ddagger$ multiple maximum values corresponding to multiple internal states.}
\label{tab:likelihoods}
\end{table}
